---
title: "class 5: Data Visualization"
author: "Joshua"
format: gfm
---

# Plotting in R

R has multipled plotting and graphics systems. The most popular of which is **ggplot2**

```{r}
plot(cars)

```

Compared to base R plots ggplot is much more verbose. 

To use ggplot one needs to first install the ggplot2 package, using the `install.packages()` command. This only needs to be done once. However, one must load the library every new session using the `library()` call.

```{r}
library(ggplot2)
```

```{r}
bp <- ggplot(cars, aes(x=speed, y = dist)) +
  geom_point()
bp

```
```{r}
bp + geom_smooth(method = "lm", se=FALSE)
```
Loading the differential expression dataset

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
ncol(genes)

table(genes$State)

100* sum(genes$State == "up") / nrow(genes) 
```

Plotting gene expression data
```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col = State) +
    geom_point()
p

```

```{r}
p + scale_colour_manual( values=c("blue","gray","red") ) + 
  labs(title = "Gene Expression Changes upon drug treatment", 
       x = "Control (No Drug)", 
       y = "Drug Treatment")

```
```{r}

```

